Reducing Horton-Strahler Stream Order Can Enhance Flood Inundation Mapping Skill with Applications for the U.S. National Water Model

### 2021-10-19 manuscripts/owp_fim3_2020/sections/results.tex
Updates * first draft complete * version sent out to JJ

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
\label{sec:results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computational Performance}
\label{ssec:compuational_performance}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
All computational experiments were conducted on an Amazon Web Services (AWS) Elastic Cloud Compute (EC2) T3 instance (2x-large size) with 4 cores, 2 threads per core, and 32GB of random access memory (RAM).
An Elastic Block Storage (EBS) solid state drive (SSD) of the general purpose 2 (gp2) type was used for 2TB of storage.
All tests took place within the project's Docker image providing the operating system and software dependencies.
Table \ref{tab:comp_results} details the computational metrics including CPU time, RAM usage, and data storage footprint for both the hydrofabric production and NRT generation of FIMs.
For the hydrofabric production, computational results were produced for the entire NWM FIM domain noted in \ref{ssec:datasets}.
For the NRT production of FIM from the hydrofabric, computational results were aggregated for the 48 HUC study area.
Computation of the depths and inundation rasters as well as the inundation in polygon form are included in this anaylsis.
Only a subset of NWM FIM domain ever gets mapped for operational purposes to avoid computational burdens by applying a 1.5\% NWM recurrence flow threshold.

\begin{table}
\caption{Descriptive computational statistics for `Cahaba' hydrofabric and NRT FIM production.}
\label{tab:comp_results}
\centering
\begin{tabular}{|p{3cm}|p{3cm}|p{3cm}|p{3cm}|}
%\hline
%"" & \multicolumn{7}{l}{CPU Time (Hrs)} & \multicolumn{7}{l}{RAM (GB)} \\
\hline
Variable & Statistic & Hydrofabric & FIM \\
\hline
\multirow{7}{*}{CPU Time (hrs)} &
n$^{a}$ & XX & XX \\ 
\cline{2-4}
& Sum  & XX & XX \\ 
\cline{2-4}
& Mean & XX & XX \\ 
\cline{2-4}
& Median & XX & XX \\ 
\cline{2-4}
& SD$^{b}$ & XX & XX \\ 
\cline{2-4}
& Max & XX & XX \\ 
\cline{2-4}
& Min & XX & XX \\ 
\hline
\multirow{7}{*}{RAM (GB)} &
n$^{a}$ & XX & XX \\
\cline{2-4}
& Sum  & XX & XX \\
\cline{2-4}
& Mean & XX & XX \\
\cline{2-4}
& Median & XX & XX \\
\cline{2-4}
& SD$^{b}$ & XX & XX \\
\cline{2-4}
& Max & XX & XX \\
\cline{2-4}
& Min & XX & XX \\
\hline
\multirow{7}{*}{Data Storage (GB)} &
n$^{a}$ & XX & XX \\
\cline{2-4}
& Sum  & XX & XX \\
\cline{2-4}
& Mean & XX & XX \\
\cline{2-4}
& Median & XX & XX \\
\cline{2-4}
& Std.Dev.$^{b}$ & XX & XX \\
\cline{2-4}
& Max & XX & XX \\
\cline{2-4}
& Min & XX & XX \\
\hline
\multicolumn{2}{l}{$^{a}$Sample size. $^{b}$Standard Deviation}
\end{tabular}
\end{table}

For comparison purposes, OWP FIM v1 measured that the entire CONUS area excepting a few HUC6's took 1.34 CPU years \cite{liu2016cybergis}. 
OWP FIM v2, developed by \citeA{djokic2019arc}, was estimated to take 0.55 CPU years for the entire NWM stream resolution but lacked the ability to run in parallel on HPC environments.
By utilizing solid programming practices and optimal algorithm selection, we were able to reduce computational requirements while enhancing hydrologic performance.

% TEST SCRIPT: 
% for huc in $(cat /data/temp/ble_test_cases_fa_20210418.lst); do echo HUC: $huc; /usr/bin/time -v /foss_fim/tools/inundation.py -r /data/outputs/first_gms_batch_test/"$huc"/rem_zeroed_masked.tif -c /data/outputs/first_gms_batch_test/"$huc"/gw_catchments_reaches_filtered_addedAttributes.tif -b /data/outputs/first_gms_batch_test/"$huc"/gw_catchments_reaches_filtered_addedAttributes_crosswalked.gpkg -t /data/outputs/first_gms_batch_test/"$huc"/hydroTable.csv -f /data/test_cases/ble_test_cases/validation_data_ble/"$huc"/100yr/ble_huc_"$huc"_flows_100yr.csv -p /data/temp/inundation_py_benchmarking_for_manuscript/inundation_"$huc".gpkg -m filter -i /data/temp/inundation_py_benchmarking_for_manuscript/inundation_"$huc".tif -d /data/temp/inundation_py_benchmarking_for_manuscript/depths_"$huc".tif -u /data/inputs/wbd/WBD_National.gpkg -l WBDHU8 -q | tee /data/temp/inundation_py_benchmarking_for_manuscript/test_1.log; done

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mapping Performance}
\label{ssec:mapping_performance}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Results}
\label{sec:results}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Mapping Performance}
\label{ssec:mapping_performance}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
We produced FIMs for the entire BLE domain within the 49 HUC8 area across several states in the south central US. 
The forecasted FIMs using the discharges for the 1\% (100 year) and 0.2\% (500 year) recurrence flows directly from HEC-RAS were used to avoid noise and errors from hydrological processes.
We computed the statistics (CSI, POD, and FAR) for both 100 and 500 year events for Mannings N set to 0.06 and 0.12. 
The distribution of these statistics can be examined in Figure \ref{fig:violin_plot} as violin plots.
Each half of a violin plot represents the kernel density estimation (KDE) for a given model (FR, MS, GMS), given Manning's N value (0.06, 0.12), and given recurrence interval (1\%, 0.2\%), and performance metric (CSI, POD, FAR).
We also denote trend lines for each metric and Manning's n setting as well as their respective slope estimate and one-tailed p-value denoting the level of significance of the trend.

Aggregating the metrics in the method above treats each HUC8 as it's own unit and does little to consider the size differences of the HUCs. 
In an opposing aggregation method, we illustrate in Table \ref{tab:aggregate_metrics} the sum of all the TPs, FPs, and FNs to recompute CSI, POD, and FAR across the entire domain. 
%
\begin{table}[h!]
\caption{Primary metrics, TPs, FPs, and FNs, aggregated for BLE domain to recompute CSI, POD, and FAR.}
\label{tab:aggregate_metrics}
\centering
%\begin{tabular}{|p{2cm}|p{2cm}|p{2cm}|p{2cm}|}
\begin{tabular}{|c|c||c|c|c|c|c|c|}
\hline
\multirow{2}{*}{Metric} & \multirow{2}{*}{Manning's N} & \multicolumn{2}{|c|}{FR} & \multicolumn{2}{|c|}{MS} & \multicolumn{2}{|c|}{GMS} \\
\cline{3-8}
  &  & 100yr & 500yr & 100yr & 500yr & 100yr & 500yr \\
\hline
\multirow{2}{*}{CSI} & 0.06 & 0.5576 & 0.5839 & 0.5717 & 0.5990 & 0.5796 & 0.6075 \\
\cline{2-8}
  & 0.12 & 0.5915 & 0.6149 & 0.6054 & 0.6288 & 0.6182 & 0.6435 \\
\hline
\multirow{2}{*}{POD} & 0.06 & 0.6354 & 0.6575 & 0.6524 & 0.6755 & 0.6633 & 0.6863 \\
\cline{2-8}
  & 0.12 & 0.7255 & 0.7446 & 0.7460 & 0.7648 & 0.7606 & 0.7810 \\
\hline
\multirow{2}{*}{FAR} & 0.06 & 0.1800 & 0.1609 & 0.1778 & 0.1589 & 0.1787 & 0.1589 \\
\cline{2-8}
  & 0.12 & 0.2379 & 0.2208 & 0.2374 & 0.2204 & 0.2324 & 0.2148 \\
\hline
\end{tabular}
\end{table}
%
\begin{figure}[h!]
\centering
\includegraphics[scale=0.9]{figures/violin_plots.jpg}
\caption{Shows kernel density estimation of the distributions (N = 49) for 1\% (100 year) and 0.2\% (500 year) along with horizontal, dashed lines for the 25th, 50th, and 75th percentiles (in order from bottom to top).
The sub-figures separate the combination of three metrics (CSI, POD, and FAR) for two settings of Manning's N (0.06 and 0.12).
Trend lines for each metric / Mannings combination are shown (N = 294) along with associated slope and p-value of slope testing one-tailed significance.}
\label{fig:violin_plot}
\end{figure}
%
% Interpretation of metrics
From Figure \ref{fig:violin_plot} and Table \ref{tab:aggregate_metrics}, we denote several meaningful trends. 
Using CSI as an overall proxy for skill of the FIM, we note that generally speaking the skill is correlated with a reduction of the stream orders of the processing units used for HAND.
In other words, the more you derive HAND on networks of unit drainage density and mosaic the resulting FIMs, the better those FIMs perform.
While this trend is evident for both sets of Manning's N values, the trend is slightly more significant for the higher value of 0.12.
Other trends related to this figure include the general performance premium for 0.2\% events as opposed to lower 1\% events.
We also note how the higher Manning's N value enhances performance for both of these recurrence intervals across all models.

Dissecting the improvements and trends presented in the previous paragraph comes down mostly to improvement in POD or a reduction in absolute amount of FNs.
POD being the primary driver in skill enhancement is evident across models by comparing the slope of the POD lines with the slope of the FAR lines.
Even though aggregating metrics by HUC8 yields a statistically zero trend, one does see a slight reduction in FAR across models that reduce HAND's maximum stream order.
Additionally, we note that POD is a primary driver in enhancing performance across Manning's N values as well.
This significant improvement comes at a cost of false alarms as the FAR increases significantly across Manning's N values.
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Computational Performance}
\label{ssec:compuational_performance}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
The NFIE experiments were able to produce HAND for 331 HUC6's in 1.34 CPU years and estimates using work from \citeA{djokic2019arc} put producing HAND at the FR NWM at 0.55 CPU years. 
For our work, we were able to produce HAND at the full NWM resolution in 0.13 CPU years which represents a substantial speed-up compared to previous works.
For the MS resolution an additional, 0.05 CPU years is required on top of this bringing the total to about 0.18 CPU years to produce 2,188 HUC8s that span additional areas not covered in previous HAND versions including Hawaii and Puerto Rico.
GMS which generalizes HAND production to level path scale adds a significant amount of CPU time to the process bringing the estimate total to about 1.17 CPU years \note[Fernando Aristizabal]{Fernando A: Review}.
%

### 2021-10-19 manuscripts/owp_fim3_2020/sections/title_authors.tex
Updates * first draft complete * version sent out to JJ

% Example: \title{This is a test title}

\title{Reducing Horton-Strahler Stream Order Can Enhance Flood Inundation Mapping Skill with Applications for the U.S. National Water Model}

%% ------------------------------------------------------------------------ %%
%
%  AUTHORS AND AFFILIATIONS
%
%% ------------------------------------------------------------------------ %%

@@

% Example: \authors{A. B. Author\affil{1}\thanks{Current address, Antartica}, B. C. Author\affil{2,3}, and D. E.
% Author\affil{3,4}\thanks{Also funded by Monsanto.}}

\authors{Fernando Aristizabal\affil{1,2,3}, Fernando Salas\affil{3}, Brian Avant\affil{1,3}, Bradford Bates\affil{1,3}, Trevor Grout\affil{1,3}, Gregory Petrochenkov\affil{4}, Nick Chadwick\affil{1,3}, Zachary Wills\affil{3,5}, Fred Ogden \affil{3}, Jasmeet Judge\affil{2}}

\affiliation{1}{Lynker Technologies, Leesburg, VA, USA}
\affiliation{2}{Center for Remote Sensing, University of Florida, Gainesville, FL, USA}
\affiliation{3}{National Water Center, Office of Water Prediction, National Oceanic and Atmospheric Administration, Tuscaloosa, AL, USA}
\affiliation{4}{New York Water Science Center, United States Geological Survey, Troy, NY, USA}
\affiliation{5}{Cooperative Institute for Satellite Earth System Studies, University of Alabama, Tuscaloosa, AL, USA}

%(repeat as many times as is necessary)

%% Corresponding Author:
% Corresponding author mailing address and e-mail address:

% (include name and email addresses of the corresponding author.  More
% than one corresponding author is allowed in this LaTeX file and for
% publication; but only one corresponding author is allowed in our
% editorial system.)

% Example: \correspondingauthor{First and Last Name}{email@address.edu}

\correspondingauthor{Fernando Aristizabal}{fernando.aristizabal@noaa.gov}

>>>>

% Example: \title{This is a test title}
%
\title{Reducing Horton-Strahler Stream Order Can Enhance Flood Inundation Mapping Skill with Applications for the U.S. National Water Model}
%
%% ------------------------------------------------------------------------ %%
%
%  AUTHORS AND AFFILIATIONS
%
%% ------------------------------------------------------------------------ %%

@@

% Example: \authors{A. B. Author\affil{1}\thanks{Current address, Antartica}, B. C. Author\affil{2,3}, and D. E.
% Author\affil{3,4}\thanks{Also funded by Monsanto.}}
%
%
\authors{Fernando Aristizabal\affil{1,2,3}, Fernando Salas\affil{3}, Gregory Petrochenkov\affil{4}, Trevor Grout\affil{1,3}, Brian Avant\affil{1,3}, Bradford Bates\affil{1,3}, Ryan Spies\affil{1,3}, Nick Chadwick\affil{1,3}, Zachary Wills\affil{3,5}, Jasmeet Judge\affil{2}}
%
%
\affiliation{1}{Lynker Technologies, Leesburg, VA, USA}
\affiliation{2}{Center for Remote Sensing, University of Florida, Gainesville, FL, USA}
\affiliation{3}{National Water Center, Office of Water Prediction, National Oceanic and Atmospheric Administration, Tuscaloosa, AL, USA}
\affiliation{4}{New York Water Science Center, Hydrologic Applied Innovations Lab, United States Geological Survey, Troy, NY, USA}
\affiliation{5}{Cooperative Institute for Satellite Earth System Studies, University of Alabama, Tuscaloosa, AL, USA}
%
%(repeat as many times as is necessary)
%
%% Corresponding Author:
% Corresponding author mailing address and e-mail address:

% (include name and email addresses of the corresponding author.  More
% than one corresponding author is allowed in this LaTeX file and for
% publication; but only one corresponding author is allowed in our
% editorial system.)
%
% Example: \correspondingauthor{First and Last Name}{email@address.edu}
%
\correspondingauthor{Fernando Aristizabal}{fernando.aristizabal@noaa.gov}
